Driver coverage

### 2024-02-28 infercnv/reports/gains_and_losses.Rmd
introducing genvisr to driver coverage plotting

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  sce: NULL
  cell_group_name: NULL
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
---

@@

```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
cnv_regions %>%
  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
               aes(y = ..density..), 
               position = position_stack(), 
               colour = NA) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
               aes(y = -..density..), 
               position = position_stack(reverse = T), 
               colour = NA) +
  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free', independent = 'x') +
  theme_bw() +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(
    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
```

@@

@@

We will define chr7+ cells as those with >80% of genes affected by duplication and chr7 WT cells as those with <20% of genes affected. We will pseudobulk by patient, cell group, and chr7 ploidy to produce a reduced object.

```{r chr7_gain, class.source = 'fold-show'}
sce$chr7_gain <-
  ifelse(sce$proportion_dupli_chr7 > 0.8, 1, 
         ifelse(sce$proportion_dupli_chr7 < 0.2, 0, 
                NA))
```

```{r chr7_gain_prop, result = 'asis'}
dat <-
  tibble::tibble(chr7_gain_value = sce$chr7_gain,
               cell_group = sce$cell_group) %>%
  dplyr::mutate(
    chr7_gain_status = dplyr::case_when(
      is.na(chr7_gain_value) ~ 'unknown/intermediate',
      chr7_gain_value == 0 ~ 'WT',
      chr7_gain_value == 1 ~ 'gain')
  )

>>>>

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  sce: NULL
  cell_group_name: NULL
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
  max_prop_for_wt: 0.5
  min_prop_for_cnv: 0.8
---

@@

```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
cnv_regions %>%
  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
               aes(y = ..density..), 
               position = position_stack(), 
               colour = NA) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
               aes(y = -..density..), 
               position = position_stack(reverse = T), 
               colour = NA) +
  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free_x', independent = 'x') +
  theme_bw() +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(
    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
```

@@

## Assigning whole chromosome-level events

Based on the observed distribution of `proportion_dupli_chr7` values and comparing this with the `inferCNV` heatmaps, whole chromosome gains are defined as `proportion_dupli_chr*` > `r params$min_prop_for_cnv`. Diploidy is defined as `proportion_dupli_chr*` < `r params$max_prop_for_wt`.

```{r assign_gains, class.source = 'fold-show'}
sce <- xfun::cache_rds({

  # apply cut-offs for whole chr gains and losses
  paste0('chr', 1:22) %>%
    purrr::walk(function(chr) {
      chr_dupli <-
        sce@colData[, paste0("proportion_dupli_", chr)]
      sce@colData[, paste0(chr, '_gain')] <<-
        ifelse(chr_dupli > params$min_prop_for_cnv, 1, 
               ifelse(chr_dupli < params$max_prop_for_wt, 0, 
                      NA))
      chr_loss <-
        sce@colData[, paste0("proportion_loss_", chr)]
      sce@colData[, paste0(chr, '_loss')] <<-
        ifelse(chr_loss > params$min_prop_for_cnv, 1, 
               ifelse(chr_loss < params$max_prop_for_wt, 0, 
                      NA))
    })

}, file = 'sce_w_gains.rds', rerun = params$rerun)
```

We also can observe the rate coincidence of chr7 and other gains. The LCM-WES data suggested a nested relationship in which there were chr5+ subclones within chr7+ populations. We look for this here.

```{r chr7_and_chr5}
events <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::select(sample, cell, cell_group, 
                dplyr::ends_with('_gain'), 
                dplyr::ends_with('_loss')) %>%
  tidyr::pivot_longer(-c('sample', 'cell', 'cell_group')) %>%
  dplyr::group_by(sample, cell, cell_group) %>%
  # remove cells in which no calls made
  dplyr::filter(!all(is.na(value))) %>%
  dplyr::mutate(
    event = dplyr::case_when(
      all(value == 0 | is.na(value)) ~ 'WT',
      value == 1 ~ name,
      TRUE ~ NA)) %>%
  dplyr::summarise(
    events = paste(sort(unique(na.omit(event))), collapse = ','),
    n_events = sum(value, na.rm = T))

p_dat <-
  events %>%
  dplyr::ungroup() %>%
  dplyr::add_count(events) %>%
  dplyr::mutate(prop = n / dplyr::n()) %>%
  # ignore events that happen in <10 cells
  dplyr::filter(events != 'WT', n >= 10) %>%
  dplyr::mutate(
    events = events %>% 
      gsub('chr', '', .) %>% 
      gsub('_gain', '+', .) %>%
      gsub('_loss', '+', .),
    events = forcats::fct_infreq(events),
    label1 = scales::label_comma()(n), 
    label = dplyr::case_when(
      n < 10 ~ label1,
      TRUE ~ paste0(label1, ' (', round(prop*100, 2), '%)')
    ))
p_dat %>%
  ggplot(aes(x = events, fill = cell_group, label = label)) +
  geom_bar() +
  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = 'bottom')
p_dat %>%
  ggplot(aes(x = forcats::fct_infreq(events), fill = sample, label = label)) +
  geom_bar() +
  geom_text(aes(y = n), hjust = 0, vjust = -0.1, angle = 45, size = 2) +
  scale_x_discrete(guide = guide_axis(angle = -90)) +
  ylim(c(0, max(p_dat$n) * 1.1)) +
  theme_classic() +
  theme(legend.position = 'none')
```

@@

We will pseudobulk by patient, cell group, and chr7 ploidy to produce a reduced object.

```{r chr7_gain_prop, results = 'asis'}
dat <-
  tibble::tibble(chr7_gain_value = sce$chr7_gain,
               cell_group = sce$cell_group) %>%
  dplyr::mutate(
    chr7_gain_status = dplyr::case_when(
      is.na(chr7_gain_value) ~ 'unknown/intermediate',
      chr7_gain_value == 0 ~ 'WT',
      chr7_gain_value == 1 ~ 'gain')
  )

### 2024-02-28 scomatic/reports/driver_coverage.Rmd
introducing genvisr to driver coverage plotting

---
title: "Driver coverage"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  id: NULL
  drivers: NULL
  ref_cds: NULL
  cache_dir: './driver_coverage_cache/'
  rerun: true
---

@@

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)

@@

We read in drivers and their depths.

```{r read_cov}
# read coverage
cov <- xfun::cache_rds({
  cov <-
    list.files(pattern = '_depths.tsv') %>% 
    purrr::set_names(., gsub('_depths.tsv', '', .)) %>%
    purrr::map(function(file) {
      file %>%
        readr::read_tsv() %>%
        dplyr::mutate(chr = as.character(chr))
    }) %>%
    dplyr::bind_rows(.id = 'id')
  cov
}, file = 'cov.rds', rerun = params$rerun)

# read drivers
drivers <- readr::read_tsv(params$drivers)
```

>>>>

---
title: "Driver coverage"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: false
    code_folding: show
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  drivers: NULL
  ref_cds: NULL
  cache_dir: './driver_coverage_cache/'
  rerun: true
---

@@

# libraries
library(magrittr)
library(ggplot2)
library(gridExtra)
library(GenomicRanges)
library(BSgenome)

@@

We read in drivers and their coverage by sample / celltype.

```{r read_cov}
# read drivers
drivers <- readr::read_tsv(params$drivers)

# read coverage
cov <- xfun::cache_rds({

  cov <-
    list.files(pattern = '_coverage.tsv') %>% 
    purrr::set_names(., gsub('_coverage.tsv', '', .)) %>%
    purrr::map(function(file) {

      # read  
      df <- 
        readr::read_tsv(file, show_col_types = F) %>%
        dplyr::rename(all = `NA`)

      # reformat for GenVisR
      df %>%
        dplyr::select(-c(chr, pos, gene)) %>%
        colnames() %>%
        purrr::set_names(., .) %>%
        purrr::map(function(ct) {
          df %>%
            dplyr::transmute(
              gene,
              chromosome = paste0('chr', chr), 
              end = pos, 
              cov = get(ct))

          }) %>%
        dplyr::bind_rows(.id = 'celltype')

    }) %>%
    dplyr::bind_rows(.id = 'id') %>%
    split(., .$gene)

}, file = 'cov.rds', rerun = params$rerun)
```

We also load the `BSgenome` and `TxDb` objects.

```{r bsgenome}
# BSgenome
library("BSgenome.Hsapiens.UCSC.hg38")
genome <- BSgenome.Hsapiens.UCSC.hg38

# TxDb
library("TxDb.Hsapiens.UCSC.hg38.knownGene")
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
```

Then we plot the coverage.

```{r gencov}
plot_cov <- function(cov, gene, var, expand_features_of_interest = T) {

  # split by variable(s)
  x <-
    cov[[gene]] %>%
    dplyr::select(chromosome, end, cov) %>%
    split(., cov[[gene]][, var])

  # make granges obj
  gr <-
    GRanges(
      seqnames = as.character(unique(cov[[gene]]$chromosome)),
      ranges = IRanges(start = min(cov[[gene]]$end) - 500,
                       end = max(cov[[gene]]$end) + 500))

  # check what to transform
  transform <- c('Intron', 'CDS', 'UTR')
  base <- c(10, 2, 2)
  if (expand_features_of_interest == T) { base <- transform <- NA }

  # plot
  GenVisR::genCov(
    x = x,
    txdb = txdb, 
    gr = gr, 
    genome = genome,
    cov_plotType = 'line',
    cov_plotLayer = ggplot2::theme(axis.title.y = ggplot2::element_blank()),
    label_txtSize = 4,
    lab2plot_ratio = c(2, length(x)),
    gene_labelTranscriptSize = 2,
    gene_name = gene,
    base = base,
    transform = transform)

}

plot_cov(cov, 'ARID2', 'celltype') 
plot_cov(cov, 'ARID2', 'celltype', F)
plot_cov(cov, 'ARID2', 'id')
plot_cov(cov, 'ARID2', 'id', F)
```

```{r exit}
knitr::knit_exit()
```
